---
output:
  word_document:
    reference_docx: "template - Copy.docx"
params:
  lake: NULL
  station: NULL
  stations: NULL
  town: NULL
  TABLE_PATH: NULL
  pub: NULL
  year: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)

# Load dotenv if available
if (file.exists(".env")) {
  if (!requireNamespace("dotenv", quietly = TRUE)) {
    install.packages("dotenv")
  }
  library(dotenv)
  load_dot_env(".env")
}

# Access environment variables
project_path <- CONFIG$PROJECT_PATH
plot_path <- CONFIG$OUTPUT_PATH
excerpt_path <- CONFIG$EXCERPT_PATH

# Load libraries
library(knitr)
library(dplyr)
library(kableExtra)
library(flextable)
library(officer)
library(magrittr)
library(officedown)

lake_clean <- tolower(gsub(" ", "_", lake))
town_clean <- tolower(gsub(" ", "_", town))

# lake-specific excerpt directory
lake_excerpt_dir <- file.path(
  excerpt_path,
  paste0(lake_clean, "_", town_clean)
)

summary_file <- file.path(
  lake_excerpt_dir,
  "summary.md"
)

recommendations_file <- file.path(
  lake_excerpt_dir,
  "recommendations.md"
)

observations_file <- file.path(
  lake_excerpt_dir,
  "observations.md"
)

read_md <- function(path) {
  if (!file.exists(path)) return("")
  paste(readLines(path, warn = FALSE), collapse = "\n")
}

```

# **`r stringr::str_to_title(lake)` – `r stringr::str_to_title(town)` (`r params$year`)**

**Water quality summary:** `r read_md(summary_file)`

**Recommended actions:** `r read_md(recommendations_file)`

## **Historical Water Quality Trend Analysis**

::: {custom-style="table caption"}
Table 1. Historical Water Quality Trends for `r stringr::str_to_title(lake)` – `r stringr::str_to_title(town)`
:::

``` {r trend table}

accent1_60 <- "#DCE6F1"  # light blue fill (body)
accent1_80 <- "#C6D9F1"  # slightly darker blue (header)

# clean and escape station id for regex
station_clean <- gsub(" ", "_", stations)
station_clean <- gsub("([\\(\\)\\[\\]\\+\\?\\*\\.\\^\\$])", "\\\\\\1", station_clean)

trend_file <- list.files(
  path = TABLE_PATH,
  pattern = paste0("^MK_TrendSummary_", station_clean, "\\.csv$"),
  full.names = TRUE,
  ignore.case = TRUE
)

if (length(trend_file) == 1) {
  
  trend_data <- read.csv(trend_file, stringsAsFactors = FALSE)
  
  # Clean column names and round numeric values
  trend_data <- trend_data |>
    rename_with(~ gsub("_", " ", .x)) |>
    mutate(across(where(is.numeric), round, 2))
  
  # Convert to flextable
  ft_trend <- flextable(trend_data) |>
    
    # styling
    bold(part = "header") |>
    bold(j = c(1, 3), part = "body") |>     # parameter columns
    
    # background colors
    bg(bg = accent1_80, part = "header") |> # darker header
    bg(bg = accent1_60, part = "body") |>   # light body
    
    # borders
    border_outer(part = "all", border = fp_border(color="black", width=1)) |>
    border_inner_h(part = "all", border = fp_border(color="black", width=1)) |>
    border_inner_v(part = "all", border = fp_border(color="black", width=1)) |>
    
    # alignment (ALL centered)
    align(align = "center", part = "all") |>
    
    # spacing & typography
    line_spacing(space = 1) |>
    padding(padding.top = 0, padding.bottom = 0, part = "all") |>
    fontsize(size = 11, part = "all") |>
    font(fontname = "Calibri", part = "all") |>
    
    # layout
    set_table_properties(width = 0.4, layout = "autofit") |>
    autofit()
  
  ft_trend
  
} else {
  cat("**Trend summary not found for", stations, "**")
}

```


## **Historical Water Quality Graphics - Deep Spot**
```{r plots, echo=FALSE, results='asis', include=FALSE}
station_id <- params$stations

plot_dirs <- c(
  file.path(plot_path, "pH_conduc"),
  file.path(plot_path, "chl_tp_secchi"),
  file.path(plot_path, "temp_DO"),
  file.path(plot_path, "plankton")
)

plot_files <- unlist(lapply(plot_dirs, function(d) {
  if (dir.exists(d)) {
    files <- list.files(
      d,
      pattern = paste0(station_id, ".*\\.(png|jpg)$"),
      full.names = TRUE,
      ignore.case = TRUE
    )
    files[file.exists(files)]
  } else {
    character(0)
  }
}), use.names = FALSE)

plot_files <- plot_files[!is.na(plot_files) & nzchar(plot_files)]
plot_files <- plot_files[seq_len(min(4, length(plot_files)))]
```

```{r trend captions, include=FALSE}
trend_files <- list.files(
  path = TABLE_PATH,
  pattern = paste0("^MK_TrendSummary_", station_clean, "\\.csv$"),
  full.names = TRUE,
  ignore.case = TRUE
)

trend_data <- NULL
trend_lookup <- NULL

if (length(trend_files) > 0) {

  # deterministic file selection
  trend_file <- trend_files[1]
  trend_data <- read.csv(trend_file, stringsAsFactors = FALSE)

  # OPTIONAL: ordering (only if column exists)
  param_order <- c(
    "Conductivity (Epilimnion)",
    "Chlorophyll-a",
    "pH (Epilimnion)",
    "Transparency",
    "Phosphorus (Epilimnion)",
    "Phosphorus (Hypolimnion)"
  )

  if ("Parameter" %in% names(trend_data)) {

    trend_data <- trend_data |>
      mutate(
        PARAM_UPPER = toupper(Parameter),
        param_index = match(PARAM_UPPER, toupper(param_order))
      ) |>
      arrange(param_index)

# --- FIX: gather BOTH column pairs into long format ---
trend_long <- trend_data |>
  dplyr::transmute(
    Parameter_1 = Parameter,
    Trend_1 = Trend,
    Parameter_2 = Parameter.1,
    Trend_2 = Trend.1
  ) |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = c(".value", "idx"),
    names_pattern = "(Parameter|Trend)_(.)"
  ) |>
  dplyr::filter(!is.na(Parameter))

# Build lookup from ALL parameters
    trend_lookup <- setNames(
      tolower(trend_long$Trend),
      trend_long$Parameter
    )
  }
}

# SAFE helper (never breaks captions)
get_trend <- function(param_name) {
  if (is.null(trend_lookup)) {
    return("no significant")
  }

  if (param_name %in% names(trend_lookup)) {
    return(trend_lookup[[param_name]])
  }

  "no significant"
}

all_no_sig <- function(params) {
  trends <- sapply(params, get_trend)
  all(trends == "no significant")
}
```

```{r fig setup, include=FALSE}
fig_counter <- 1
plot1_exists <- length(plot_files) >= 1
plot2_exists <- length(plot_files) >= 2
plot3_exists <- length(plot_files) >= 3
plot4_exists <- length(plot_files) >= 4
```

```{r cond pH plot, out.width = '90%'}
if (plot1_exists) {
  knitr::include_graphics(plot_files[1])
}
```

`r if (plot1_exists) {

paste0(
"::: {custom-style=\"figure caption\"}\n",

if (all_no_sig(c("pH (Epilimnion)", "Conductivity (Epilimnion)"))) {

paste0(
"Figure ", fig_counter,
". Median epilimnetic conductivity (red points) and pH (gray bars) by year."
)

} else {

paste0(
"Figure ", fig_counter,
". Median epilimnetic conductivity (red points) and pH (gray bars) by year, ",
"with corresponding trend lines shown as red and black dashed lines, respectively. ",
"Epilimnion pH is ", get_trend("pH (Epilimnion)"),
" and conductivity is ", get_trend("Conductivity (Epilimnion)"),
" since monitoring began."
)

},

"\n:::"
)

} else {

""  # no caption at all
}`

```{r cond pH plot fig counter, out.width = '90%'}
if (plot1_exists) fig_counter <- fig_counter + 1
```

```{r chl tp trans plot, out.width = '90%'}
if (plot2_exists) {
  knitr::include_graphics(plot_files[2])
}
```

`r if (plot2_exists) {

paste0(
"::: {custom-style=\"figure caption\"}\n",

if (all_no_sig(c(
"Transparency",
"Phosphorus (Epilimnion)",
"Chlorophyll-a"
))) {

paste0(
"Figure ", fig_counter,
". Median Secchi disk transparency (blue bars), epilimnetic phosphorus (red triangles), ",
"and chlorophyll-a (green points) by year."
)

} else {

paste0(
"Figure ", fig_counter,
". Median Secchi disk transparency (blue bars), epilimnetic phosphorus (red triangles), ",
"and chlorophyll-a (green points) by year, with corresponding trend lines shown as blue, red, ",
"and green dashed lines, respectively. Water transparency is ",
get_trend("Transparency"), ", phosphorus is ",
get_trend("Phosphorus (Epilimnion)"), " and chlorophyll-a is ",
get_trend("Chlorophyll-a"),
" since monitoring began."
)

},

"\n:::"
)

} else {

""
}`

```{r tp chl fig counter, include=FALSE}
if (plot2_exists) fig_counter <- fig_counter + 1
```


```{r temp DO plot, out.width = '90%'}
if (plot3_exists) {
  knitr::include_graphics(plot_files[3])
}
```

`r if (plot3_exists) {

paste0(
"::: {custom-style=\"figure caption\"}\n",

"Figure ", fig_counter,
". ", params$year,
" vertical profiles of temperature (°C; green points, left plot) and dissolved oxygen (mg/L; blue points, right plot) plotted against depth (m). ",
"Profile month is indicated by point shape and color intensity, with lighter shades representing earlier-season samples and darker shades representing later-season samples.",

"\n:::"
)

} else {

""
}`

```{r temp DO fig counter, include=FALSE}
if (plot3_exists) fig_counter <- fig_counter + 1
```


```{r plankton, out.width = '90%'}
if (plot4_exists) {
  knitr::include_graphics(plot_files[4])
}
```

`r if (plot4_exists) {

paste0(
"::: {custom-style=\"figure caption\"}\n",

"Figure ", fig_counter,
". Phytoplankton community composition over time, expressed as the average relative abundance of major taxonomic groups per year.",

"\n:::"
)

} else {

""
}`

```{r plankton fig counter, include=FALSE}
if (plot4_exists) fig_counter <- fig_counter + 1
```

::: {custom-style="table caption"}
Table 2. `r params$year` Average Water Quality Data for `r stringr::str_to_title(lake)` – `r stringr::str_to_title(town)`
:::

```{r avg water quality}
lake <- params$lake
town <- params$town   # NEW: town passed in from params
accent1_60 <- "#DCE6F1"
accent1_80 <- "#C6D9F1"  # darker header

# Clean lake & town strings for file matching
lake_clean <- gsub(" ", "_", lake)
town_clean <- gsub(" ", "_", town)

# Escape regex special characters
lake_clean <- gsub("([\\(\\)\\[\\]\\+\\?\\*\\.\\^\\$])", "\\\\\\1", lake_clean)
town_clean <- gsub("([\\(\\)\\[\\]\\+\\?\\*\\.\\^\\$])", "\\\\\\1", town_clean)

# Search pattern: Lake_Town_CYA.csv
pattern <- paste0("^CYA_", lake_clean, "_", town_clean, "\\.csv$")

cya_file <- list.files(
  TABLE_PATH,
  pattern = pattern,
  full.names = TRUE,
  ignore.case = TRUE
)

if (length(cya_file) == 1) {

  cya_data <- read.csv(cya_file, check.names = FALSE, stringsAsFactors = FALSE)
  
  if ("STATNAME" %in% names(cya_data)) {
    names(cya_data)[names(cya_data) == "STATNAME"] <- "Station"
  }
  
  cya_data <- cya_data |>
    select(where(~ !all(is.na(.x) | .x == ""))) |>
    mutate(across(where(is.numeric), round, 2))
  
  ft_cya <- flextable(cya_data) |>
    bold(part = "header") |>
    bg(bg = accent1_80, part = "header") |>   # darker header
    bg(bg = accent1_60, part = "body") |>     # light body
    border_outer(part = "all", border = fp_border(color = "black", width = 1)) |>
    border_inner_h(part = "all", border = fp_border(color = "black", width = 1)) |>
    border_inner_v(part = "all", border = fp_border(color = "black", width = 1)) |>
    align(align = "center", part = "header") |>
    line_spacing(space = 1) |> 
    padding(padding.top = 0, padding.bottom = 0, part = "all") |> 
    fontsize(size = 11.0, part = "all") |>          
    set_table_properties(layout = "autofit") |>
    font(fontname = "Calibri", part = "all")
  
  if (nrow(cya_data) >= 2) {
    ft_cya <- align(ft_cya, j = 1, i = 2:nrow(cya_data), align = "left", part = "body")
  }
  if (ncol(cya_data) >= 2) {
    ft_cya <- align(ft_cya, j = 2:ncol(cya_data), align = "center", part = "body")
  }
  
  ft_cya

} else if (length(cya_file) == 0) {

  cat("**CYA summary not found for ", lake, ", ", town, "**")

} else {

  cat("**Multiple CYA files found for ", lake, ", ", town, " – please check directory**")

}
```

*Note: “No Value” indicates that the parameter was not measured at this station during the reporting period.*

## **Observations** *(Refer to Table 2 and Historical Deep Spot Data Graphics)*:

`r read_md(observations_file)` 

### **How does your lake compare to New Hampshire lakes and water quality standards?**

::: {custom-style="table caption"}
Table 3. New Hampshire Median Lake Water Quality Values. Median values generated from historic lake monitoring data.
:::

``` {r NH median values, echo = FALSE, warning = FALSE}
accent1_60 <- "#DCE6F1"
accent1_80 <- "#C6D9F1"  # darker header

# --- NH Median Lake Water Quality Values ---
median_values <- data.frame(
  Parameter = c("Alkalinity", "Chlorophyll-a", "Chloride", "Conductivity", 
                "Total Phosphorus", "Transparency", "pH"),
  `Median Value` = c("4.5 mg/L", "4.39 μg/L", "5 mg/L", "42.3 μS/cm",
                    "11 μg/L", "3.3 m", "6.6"),
  check.names = FALSE
)

ft_median <- flextable(median_values) |>
  bold(part = "header") |>
  bold(j = 1, part = "body") |>

  # background styling (split header vs body)
  bg(bg = accent1_80, part = "header") |>
  bg(bg = accent1_60, part = "body") |>

  border_outer(part = "all", border = fp_border(color="black", width=1)) |>
  border_inner_h(part = "all", border = fp_border(color="black", width=1)) |>
  border_inner_v(part = "all", border = fp_border(color="black", width=1)) |>

  align(align = "center", part = "header") |>

  line_spacing(space = 1) |> 
  padding(padding.top = 0, padding.bottom = 0, part = "all") |> 
  fontsize(size = 11, part = "all") |>
  set_table_properties(width = 0.35, layout = "autofit") |>
  font(fontname = "Calibri", part = "all") |>
  autofit()

ft_median

```


::: {custom-style="table caption"}
Table 4. New Hampshire Water Quality Standards. Numeric criteria for specific parameters. Water quality violation occurs if thresholds are exceeded.
:::

``` {r NH water quality standards}
# --- NH Water Quality Standards ---
accent1_60 <- "#DCE6F1"
accent1_80 <- "#C6D9F1"  # darker header

water_quality_standards <- data.frame(
  Parameter = c("Chloride", "E. coli (beach)", "E. coli (surface water)", 
                "pH", "Turbidity"),
  Threshold = c("> 230 mg/L (chronic)", "> 88 cts/100 mL", "> 406 cts/100 mL",
                "between 6.5-8.0 (unless naturally occurring)", 
                "> 10 NTU above natural"),
  check.names = FALSE
)

ft_standards <- flextable(water_quality_standards) |>
  bold(part = "header") |>
  bold(j = 1, part = "body") |>

  # background styling split
  bg(bg = accent1_80, part = "header") |>
  bg(bg = accent1_60, part = "body") |>

  border_outer(part = "all", border = fp_border(color="black", width=1)) |>
  border_inner_h(part = "all", border = fp_border(color="black", width=1)) |>
  border_inner_v(part = "all", border = fp_border(color="black", width=1)) |>

  align(align = "center", part = "header") |>

  line_spacing(space = 1) |> 
  padding(padding.top = 0, padding.bottom = 0, part = "all") |> 
  fontsize(size = 11, part = "all") |> 
  set_table_properties(width = 0.5, layout = "autofit") |>
  font(fontname = "Calibri", part = "all") |>
  autofit()

ft_standards
```